The role of aneuploidy in the evolution of cancer drug resistance

### 2023-09-05 EvolutionaryRescue1.tex
Remus

\usepackage{amssymb,amsmath,amsthm}
\usepackage{commath}
\usepackage[hyphens]{url}
\usepackage[dvipsnames]{xcolor}
\usepackage[unicode=true,colorlinks=true,urlcolor=CadetBlue,citecolor=black,linkcolor=black]{hyperref}
\PassOptionsToPackage{hyphens}{url} % url is loaded by hyperref
\usepackage{authblk}
\usepackage{longtable}
\usepackage{multirow}
\usepackage{booktabs}
\usepackage{lipsum}  
\usepackage[title,page]{appendix}
\usepackage{chngcntr}
%\usepackage{end float}
\usepackage{cleveref}
 \usepackage{subcaption}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
%Evolutionary rescue is the process by which a population is able to survive a sudden environmental change which initially causes the population to decline towards extinction. A prime example of evolutionary rescue is the ability of cancer to survive being exposed to various treatments. We are interested in the mechanisms through which a population of cancer cells are able to adapt to chemotherapy, and in particular, the role played by chromosomal instability (aneuploidy). Cancer cells which have aneuploidy are hypothesized to have a higher fitness in an environment altered by anti-cancer drugs as they have incomplete pathways which drugs activate in order to kill the cells. Aneuploidy is highly prevalent in tumors and certain drugs which attempt to combat cancers through increasing chromosomal instability. As a result, the question we wish to answer is how aneuploidy impacts the fate of the population of cancer cells. We propose to model evolutionary rescue with the help of multi-type branching processes to obtain the probability that cancer will survive. Additionally, we will utilize large genomic datasets to asses the effects of aneuploidy on the probability of evolutionary rescue.
\end{abstract}

@@

% TODO two papers that show aneuploidy provides resistance
% Lukow, Devon A., Erin L. Sausville, Pavit Suri, Narendra Kumar Chunduri, Angela Wieland, Justin Leu, Joan C. Smith, et al. 2021. “Chromosomal Instability Accelerates the Evolution of Resistance to Anti-Cancer Therapies.” Developmental Cell 56 (17): 2427-2439.e4. https://doi.org/10.1016/j.devcel.2021.07.009.
% Rutledge, Samuel D., Temple A. Douglas, Joshua M. Nicholson, Maria Vila-Casadesús, Courtney L. Kantzler, Darawalee Wangsa, Monika Barroso-Vilares, Shiv D. Kale, Elsa Logarinho, and Daniela Cimini. 2016. “Selective Advantage of Trisomic Human Cells Cultured in Non-Standard Conditions.” Scientific Reports 6 (June 2015): 1–12. https://doi.org/10.1038/srep22828.

\paragraph{Aneuploidy in cancer.} Chromosomal instability (CIN) is the mitotic process in which cells suffer from chromosome mis-segregation that leads to aneuploidy, where cells are characterized by structural changes of the chromosomes and copy number alterations \citep{schukken2018cin}.
Interestingly, aberrations in chromosome copy number have been shown to allow cancer cells to survive under stressful conditions such as drug therapy.
Indeed, cancer cells are often likely to be aneuploid, and aneuploidy is associated with poor patient outcomes \citep{ben2020context}.

@@

In our analysis we assume that lineages produced by cells from the initial population divide and die independently of each other, which may be unrealistic, as cells usually compete for resources.
A more realistic model includes competition for limited resources and spatial structure, which may play an important role in the development of cancer \citep[e.g.,][]{martens2011spatial}.
To simulate birth and death rates that depend on the number of cells in the population, we transform the rates of division and death to the following:
\begin{align*}
\lambda_w' &= \lambda_w, \\
\mu_w' &= \mu_w,\\
\lambda_a' &= C_1+\left(\lambda_a-\mu_a\right)\left(1-\frac{w+a+m}{K}\right),\\ 
\mu_a' &= C_1,\\
\lambda_m' &= C_2+\left(\lambda_m-\mu_m\right)\left(1-\frac{w+a+m}{K}\right),\\ 
\mu_m' &= C_2,
\end{align*}
where $C_1, C_2>0$ are constants and $K$ is the maximum carrying capcity. % TODO what constants? see Remus version.

@@

Otherwise, aneuploidy is frequent enough ($u > \max{(-\Delta_a, 1/T^*)}$) to affect the evolution of drug resistance. 
The threshold tumor size, $N^*$, can then be approximated by one of the following cases, depending on $\Delta_a T^*$, the change in the log of the aneuploid population size during the critical time,
\begin{equation}  \label{eq:N_a}
\begin{aligned}
N_a^* \approx 
  \frac{\abs{\Delta_w}}{u} \cdot \begin{cases}
    \frac{\abs{\Delta_a}}{v} \cdot \frac{\lambda_m}{\Delta_m} ,&
  \Delta_a T^* \ll -1 \text{ (partially sensitive aneuploids)},\\ 
  %\left(\frac{\lambda_a}{v} \cdot \frac{\lambda_m}{\Delta_m}\right)^{1/2} ,&
  2\lambda_a T^* ,&
  -1 \ll \Delta_a T^* \ll 1  \text{ (stationary aneuploids)},\\ 
  \frac{\lambda_a}{\Delta_a} ,&
   \Delta_a T^* \gg 1 \text{ (resistant aneuploids)}.
  \end{cases}
\end{aligned}
\end{equation}
%The second case has a finer approximation by $N^*_a \approx \frac{\abs{\Delta_w}}{u} \cdot \frac{2\lambda_a}{\Delta_a + 1/T^*}$.
The first line describes the case in which aneuploids are still effectively killed by the treatment, but not as quickly as the wild type. 
In the second case, the aneuploids are sufficiently resistant that the size of each aneuploid lineage is expected to remain roughly constant. 
In both of these first two cases, aneuploidy increases the probability of rescue by slowing or halting the decrease of the cancer population, allowing more opportunities for producing resistant mutants. 
In the third case, aneuploidy provides sufficient resistance for the aneuploid population to re-grow the tumor even without additional resistance mutations.
Note that in this case there is no dependence on the parameters characterizing mutants or their production ($v$, $\lambda_m$, and $\Delta_m$).
Comparing these approximations to results of stochastic evolutionary simulations, we find that the approximations perform very well (\Cref{rescue_prob_wt_growth,rescue_prob_an_growth,rescue_prob_N,rescue_prob_N}).

@@

In the above we assumed that at the onset of drug treatment, the initial tumor consisted entirely of wildtype cells that are drug sensitive.
However, aneuploid cells are likely generated even before onset of treatment at some rate $\tilde{u} \le u$ (because the treatment itself may promote generation of aneuploid cells REF), which are likely to have a deleterious effect in the absence of the drug, $s$ (REF). % TODO refs
But if the number of cells in the tumor $N$ is large (as expected if the tumor is to be treated with a drug), there may already be a fraction $f \approx \tilde{u}/s$ of aneuploid cells in the population.

Therefore, the threshold tumor size with standing generation variation, $\tilde{N}^*_{a}$, is similar to the ratio with de-novo variation, $N^*_a$, except that the sensitive growth rate $\abs{\Delta_w}$ is replaced with the aneuploidy cost, $s$, such that 
\begin{equation}
\frac{\tilde{N}^*_{a}}{N^*_{a}} = \frac{u}{\tilde{u}} \; \frac{s}{\abs{\Delta_w}} .
\end{equation}

@@

TODO

% ? Distribution of recurrence time -- with and without aneuploidy

@@

We find that aneuploidy can have a significant effect on evolutionary rescue (\Cref{rescue_ratio,MeanTimeRatioInitialPopulationSize}). For example, when aneuploidy cells are "barely-resistant" (they grow at a very low rate, $\Delta_a=10^{-3}$) the probability of evolutionary rescue is 1,000-fold higher with aneuploidy than without it (for parameters previously described in cancer, see \Cref{table1}).
Interestingly, aneuploidy is unlikely to contribute to evolutionary rescue in primary tumors in which the number of cells is large enough ($N>10^7$) for the appearance of resistant mutation directly in sensitive cells before these cells become extinct (\Cref{rescue_ratio}).
However, aneuploidy can have a crucial role in evolutionary rescue of secondary tumors, in which the number of sensitive cells may be below the detection threshold of $\sim10^7$  \citep{bozic2013evolutionary}.
Given the fact that the mean time for such secondary tumors to overcome chemotherapy can be of the order of 100 days (\Cref{MeanTimeGrowthAneuploidyPlot}), % TODO do we see tau=100 here...?
this can explain the reappearance of cancer even after initial remission.
Indeed, we find that the tumor size can decrease by orders of magnitude before it is rescued (\Cref{MinTumorSize}).

We hypothesized that presence of \emph{standing variation} % TODO ref
--the existence of a subpopulation of aneuploid cancer cells before therapy begins--can facilitate evolutionary rescue by reducing the waiting time for the appearance of aneuploid cells. Indeed, we observe that even when a small fraction of the initial tumor is aneuploid, evolutionary rescue is more likely to occur through this existing standing variation, rather then through \emph{de novo} aneuploid cells (\Cref{FractionPlot}).

We have assumed that cancer cell lineages are independent of each other. However, this may not be the case, as cancer cells compete for resources (e.g., blood supply). Nevertheless, we find that when the carrying capacity if large % TODO HOW LARGE?
our approximation for the probability of evolutionary rescue agrees with results of stochastic simulations with density-dependent growth  \Cref{rescue_prob_N}).
Future work may focus on scenarios with small carrying capacity by analyzing density-dependent branching processes. % TODO ref?

Our model predictions may be tested by experiments \citep{martin2013probability}. For example, to study the effects of initial tumor size on the probability of evolutionary rescue, a large culture mass can be propagated from a single cancer cell in permissive conditions and then diluted to a  range of starting tumor sizes. Afterwards, these tumors may be exposed to anti-cancer drugs that induces aneuploidy % TODO ref
or to saline solution for control. 
Cell density can then be measured and compared to the predictions of our model. % TODO not clear what is to be compared... what does our model predict...?

% TODO discussion of mean time?

% TODO summary

@@

\begin{table}
\begin{center}
  \begin{tabular}{| l |p{5cm}| c | c | p{3cm} |}
    \hline
     & Name & Value & Units & References \\ \hline
    $N$ & Initial tumor size & $10^7-10^9$ & cells  & \citet{del2009does} \\ \hline
    $\lambda_w$ & Wildtype division rate& 0.14 & 1/days  & \citet{bozic2013evolutionary} \\ \hline
    $\mu_w$ & Wildtype death rate& 0.17 & 1/days  & \citet{bozic2013evolutionary} \\ \hline
    $\lambda_a$  & Aneuploid division rate$^\ast$ & 0.14 & 1/days  & - \\ \hline
    $\mu_a$ & Aneuploid death rate$^\ast$ & $0.13-0.17$ & 1/days  & - \\ \hline
    $\lambda_m$ & Mutant division rate& 0.14 & 1/days  & \citet{bozic2013evolutionary} \\ \hline
    $\mu_m$ & Mutant death rate& 0.13 & 1/days  & \citet{bozic2013evolutionary} \\ \hline
    $u$ & Missegregation rate& $10^{-3}-10^{-2}$ & 1$\slash$cell division  & \citet{nowak2004evolutionary,bakker2023predicting} \\ \hline
    $v$ & Mutation rate& $10^{-7}-10^{-9}$ &  1$\slash$gene$\slash$cell division  & \citet{nowak2004evolutionary} \\ 
    % TODO citing Nowak 2004 for the rates is strange because it is a theory paper - cite the original empirical papers
    \hline
  \end{tabular}
\caption{\textbf{Model parameters.} NEED DIFFERENT REFS---THESE ARE MOSTLY THEORY PAPERS. CHECK BIONUMBERS?
Aneuploid birth rate $\lambda_a$ is set to the same value as the wildtype and mutant birth rates, $\lambda_w$ and $\lambda_m$.
Aneuploid death rate $\mu_a$ is set to an intermediate value between the wildtype and mutant death rates, $\mu_w$ and $\mu_m$.}
  \label{table1}
\end{center}
\end{table}

@@

\begin{figure}
\centering
\includegraphics[width=0.75\textwidth]{Figures/figureAneuploidy.pdf}
\caption{
\textbf{Model illustration.}
\textbf{(A)} A population of cancer cells is composed of wildtype, aneuploid, and mutant cells, which divide with rates $\lambda_w$, $\lambda_a$, and $\lambda_m$ and die at rates $\mu_w$, $\mu_a$, and $\mu_m$, respectively. 
Wildtype cells can become aneuploid at rate $u$. Both aneuploid and wildtype cells can acquire a beneficial mutation with rate $v$. Color denotes the relative growth rates of the three genotypes such that $\lambda_w - \mu_w < \lambda_a - \mu_a < \lambda_m - \mu_m$. \textbf{(B)} The wildtype and the mutant are susceptible and resistant, respectively, to the drug. The aneuploid may be tolerant, non-growing, or or partially resistant.
}
\label{figureAneuploidy}
\end{figure}

@@

\begin{figure}
\begin{subfigure}{0.5\textwidth}
A\\
\includegraphics[width=1\textwidth]{Figures/ProbvNPlot.pdf}
\end{subfigure}
\\
\begin{subfigure}{0.5\textwidth}
B\\
\includegraphics[width=1\textwidth]{Figures/ThresholdPopulationSizePlot.pdf}
\end{subfigure}
\begin{subfigure}{0.5\textwidth}
C\\
\includegraphics[width=1\textwidth]{Figures/ThresholdPopulationSizeVersusRatioPlot.pdf}
\end{subfigure}
\caption{
\textbf{Aneuploidy facilitates evolutionary rescue of cancer under drug treatment.}
\textbf{(A)} The probability of evolutionary rescue (i.e. the probability that the population does not go to extinction), $\presc$, as a function of the initial tumor size, $N$. Dashed line shows the threshold tumor size, $N^*$, above which the probability is very high.
\textbf{(B)} The threshold tumor size $N_a^*$ as a function of the aneuploid growth rate $\Delta_a$. The dashed horizontal line shows $N^*_m$, the threshold tumor size without aneuploidy ($u=0$). When aneuploid growth rate is close to or higher than zero, aneuploidy decreases the threshold tumor size, thereby facilitating evolutionary rescue.
\textbf{(C)} The threshold tumor size $N_a^*$ as a function of the ratio of aneuploidy and mutation rates, $u/v$. The dashed horizontal line shows $N^*_m$, the threshold tumor size without aneuploidy ($u=0$).  When the aneuploidy rate is much higher than the mutation rate, aneuploidy decreases the threshold tumor size, thereby facilitating evolutionary rescue.
}
\label{fig:LABEL}
\end{figure}
%%%%%%%%%%%

@@

@@

\section*{Appendix C: Evolutionary rescue time}
% TODO remove the linear approximation, it is not good.

>>>>

\usepackage{amssymb,amsmath,amsthm}
\usepackage{commath}
\usepackage[hyphens]{url}
\usepackage[dvipsnames]{xcolor}
\usepackage[unicode=true,colorlinks=true,urlcolor=CadetBlue,citecolor=black,linkcolor=black]{hyperref}
\PassOptionsToPackage{hyphens}{url} % url is loaded by hyperref
\usepackage{authblk}
\usepackage{longtable}
\usepackage{multirow}
\usepackage{booktabs}
\usepackage{lipsum}  
\usepackage[title,page]{appendix}
\usepackage{chngcntr}
%\usepackage{end float}
 \usepackage{subcaption}
 \usepackage{cleveref}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
Evolutionary rescue is the process by which a population is able to survive a sudden environmental change which initially causes the population to decline towards extinction. A prime example of evolutionary rescue is the ability of cancer to survive being exposed to various treatments. We are interested in the mechanisms through which a population of cancer cells are able to adapt to chemotherapy, and in particular, the role played by chromosomal instability (aneuploidy). Cancer cells which have aneuploidy are hypothesized to have a higher fitness in an environment altered by anti-cancer drugs as they have incomplete pathways which drugs activate in order to kill the cells. Aneuploidy is highly prevalent in tumors and certain drugs which attempt to combat cancers through increasing chromosomal instability. As a result, the question we wish to answer is how aneuploidy impacts the fate of the population of cancer cells. We propose to model evolutionary rescue with the help of multi-type branching processes to obtain the probability that cancer will survive. 
\end{abstract}

@@

% TODO two papers that show aneuploidy provides resistance Ok
% Lukow, Devon A., Erin L. Sausville, Pavit Suri, Narendra Kumar Chunduri, Angela Wieland, Justin Leu, Joan C. Smith, et al. 2021. “Chromosomal Instability Accelerates the Evolution of Resistance to Anti-Cancer Therapies.” Developmental Cell 56 (17): 2427-2439.e4. https://doi.org/10.1016/j.devcel.2021.07.009.
% Rutledge, Samuel D., Temple A. Douglas, Joshua M. Nicholson, Maria Vila-Casadesús, Courtney L. Kantzler, Darawalee Wangsa, Monika Barroso-Vilares, Shiv D. Kale, Elsa Logarinho, and Daniela Cimini. 2016. “Selective Advantage of Trisomic Human Cells Cultured in Non-Standard Conditions.” Scientific Reports 6 (June 2015): 1–12. https://doi.org/10.1038/srep22828.

\paragraph{Aneuploidy in cancer.} Chromosomal instability (CIN) is the mitotic process in which cells suffer from chromosome mis-segregation that leads to aneuploidy, where cells are characterized by structural changes of the chromosomes and copy number alterations \citep{schukken2018cin}.
Interestingly, aberrations in chromosome copy number have been shown to allow cancer cells to survive under stressful conditions such as drug therapy \citep{lukow2021chromosomal,rutledge2016selective}.
Indeed, cancer cells are often likely to be aneuploid, and aneuploidy is associated with poor patient outcomes \citep{ben2020context}.

@@

In our analysis we assume that lineages produced by cells from the initial population divide and die independently of each other, which may be unrealistic, as cells usually compete for resources.
A more realistic model includes competition for limited resources and spatial structure, which may play an important role in the development of cancer \citep[e.g.,][]{martens2011spatial}.
To simulate birth and death rates that depend on the number of cells in the population, we transform the rates of division and death to the following:
\begin{align*}
\lambda_w' &= \lambda_w, \\
\mu_w' &= \mu_w,\\
\lambda_a' &= \lambda_a,\\ 
\mu_a' &= \mu_a + \lambda_a\frac{w+a+m}{K},\\
\lambda_m' &= \lambda_m,\\ 
\mu_m' &= \mu_m + \lambda_m\frac{w+a+m}{K}.
\end{align*}
where $K$ is the maximum carrying capcity. % TODO what constants? see Remus version. Ok

@@

Otherwise, aneuploidy is frequent enough ($u > \max{(-\Delta_a, 1/T^*)}$) to affect the evolution of drug resistance. 
The threshold tumor size, $N^*$, can then be approximated by one of the following cases, depending on $\Delta_a T^*$, the change in the log of the aneuploid population size during the critical time,
\begin{equation}  \label{eq:N_a}
\begin{aligned}
N_a^* \approx 
  \frac{\abs{\Delta_w}}{u} \cdot \begin{cases}
    \frac{\abs{\Delta_a}}{v} \cdot \frac{\lambda_m}{\Delta_m} ,&
  \Delta_a T^* \ll -1 \text{ (partially sensitive aneuploids)},\\ 
  %\left(\frac{\lambda_a}{v} \cdot \frac{\lambda_m}{\Delta_m}\right)^{1/2} ,&
  2\lambda_a T^* ,&
  -1 \ll \Delta_a T^* \ll 1  \text{ (stationary aneuploids)},\\ 
  \frac{\lambda_a}{\Delta_a} ,&
   \Delta_a T^* \gg 1 \text{ (resistant aneuploids)}.
  \end{cases}
\end{aligned}
\end{equation}
%The second case has a finer approximation by $N^*_a \approx \frac{\abs{\Delta_w}}{u} \cdot \frac{2\lambda_a}{\Delta_a + 1/T^*}$.
The first line describes the case in which aneuploids are still effectively killed by the treatment, but not as quickly as the wild type. 
In the second case, the aneuploids are sufficiently resistant that the size of each aneuploid lineage is expected to remain roughly constant. 
In both of these first two cases, aneuploidy increases the probability of rescue by slowing or halting the decrease of the cancer population, allowing more opportunities for producing resistant mutants. 
In the third case, aneuploidy provides sufficient resistance for the aneuploid population to re-grow the tumor even without additional resistance mutations.
Note that in this case there is no dependence on the parameters characterizing mutants or their production ($v$, $\lambda_m$, and $\Delta_m$).
Comparing these approximations to results of stochastic evolutionary simulations, we find that the approximations perform very well (\Cref{rescue_prob}).%rescue_prob_wt_growth,rescue_prob_an_growth,rescue_prob_N,rescue_prob_N}).

@@

In the above we assumed that at the onset of drug treatment, the initial tumor consisted entirely of wildtype cells that are drug sensitive.
However, aneuploid cells are likely generated even before onset of treatment at some rate $\tilde{u} \le u$ (because the treatment itself may promote generation of aneuploid cells \citep{wang2019molecular,mason2017functional}), which are likely to have a deleterious effect in the absence of the drug, $s$ \citep{replogle2020aneuploidy,giam2015aneuploidy}. % TODO refs OK
But if the number of cells in the tumor $N$ is large (as expected if the tumor is to be treated with a drug), there may already be a fraction $f \approx \tilde{u}/s$ of aneuploid cells in the population.

Therefore, the threshold tumor size with standing generation variation, $\tilde{N}^*_{a}$, is similar to the ratio with de-novo variation, $N^*_a$, except that the sensitive growth rate $\abs{\Delta_w}$ is replaced with the aneuploidy cost, $s$, such that 
\begin{equation}
\frac{\tilde{N}^*_{a}}{N^*_{a}} = \frac{u}{\tilde{u}} \; \frac{s}{\abs{\Delta_w}}.
\end{equation}

@@

In Appendix D we approximate the mean time for the population of mutant cancer cells to reach the initial population size $N$. \Cref{MeanTimeGrowthAneuploidyPlot} shows the agreement between these approximations and simulation results for intermediate and large tumor sizes.

\paragraph{Distribution of recurrence time -- with and without aneuploidy.}

In Appendix E we derive the probability that by time $t$ a succesful mutant has been generated. \Cref{ReboundProbability} show the agreement between our formula and simulation results for the case when aneuploidy is present and when it is absent. 
% ? Distribution of recurrence time -- with and without aneuploidy

@@

We find that aneuploidy can have a significant effect on evolutionary rescue (\Cref{rescue_prob}). For example, when aneuploidy cells are "barely-resistant" (they grow at a very low rate, $\Delta_a=10^{-3}$) the probability of evolutionary rescue is 1,000-fold higher with aneuploidy than without it (for parameters previously described in cancer, see \Cref{table1}).
Interestingly, aneuploidy is unlikely to contribute to evolutionary rescue in primary tumors in which the number of cells is large enough ($N>10^7$) for the appearance of resistant mutation directly in sensitive cells before these cells become extinct (\Cref{rescue_ratio}).
However, aneuploidy can have a crucial role in evolutionary rescue of secondary tumors, in which the number of sensitive cells may be below the detection threshold of $\sim10^7$  \citep{bozic2013evolutionary}.
Given the fact that the mean time for such secondary tumors to overcome chemotherapy can be of the order of 100 days (\Cref{MeanTimeGrowthAneuploidyPlot}), % TODO do we see tau=100 here...?
this can explain the reappearance of cancer even after initial remission.
Indeed, we find that the tumor size can decrease by orders of magnitude before it is rescued (\Cref{MinTumorSize}).

We hypothesized that presence of \emph{standing variation} % TODO ref <- why ref?
--the existence of a subpopulation of aneuploid cancer cells before therapy begins--can facilitate evolutionary rescue by reducing the waiting time for the appearance of aneuploid cells. Indeed, we observe that even when a small fraction of the initial tumor is aneuploid, evolutionary rescue is more likely to occur through this existing standing variation, rather then through \emph{de novo} aneuploid cells (\Cref{FractionPlot}).

We have assumed that cancer cell lineages are independent of each other. However, this may not be the case, as cancer cells compete for resources (e.g., blood supply). Nevertheless, we find that when the carrying capacity is large ($K\sim10^7$) % TODO HOW LARGE? ok
our approximation for the probability of evolutionary rescue agrees with results of stochastic simulations with density-dependent growth  \Cref{rescue_prob_N}).
Future work may focus on scenarios with small carrying capacity by analyzing density-dependent branching processes~\citep{klebaner1997population}. % TODO ref? OK

Our model predictions may be tested by experiments \citep{martin2013probability}. For example, to study the effects of initial tumor size on the probability of evolutionary rescue, a large culture mass can be propagated from a single cancer cell in permissive conditions and then diluted to a  range of starting tumor sizes. Afterwards, these tumors may be exposed to anti-cancer drugs that induces aneuploidy % TODO ref OK
or to saline solution for control~\citep{ippolito2021gene}. 
Cell density can then be measured and compared to the predictions of our model. % TODO not clear what is to be compared... what does our model predict...?

We observe that the presence of aneuploidy has the effect of shortening the time necessary for a mutant to appear which will rescue the tumor (\Cref{MeanTimeGrowthAneuploidyPlot}). However, this effect is only true for small tumors ($N<10^5$) as direct mutation is the main mechanism for generation of successful mutants for large and intermediate population sizes.
% TODO discussion of mean time? ok

Our study quantitatively confirms that aneuploidy plays an important role in tumors overcoming exposure to chemotherapeutic drugs when the tumor size is small or intermediate. Very large tumors can escape anti-cancer drugs through direct mutation while smaller ones are able to obtain the beneficial mutaion through an aneuploid intermediary (\Cref{rescue_prob}).
% TODO summary ok

@@

\begin{table}
\begin{center}
  \begin{tabular}{| l |p{5cm}| c | c | p{3cm} |}
    \hline
     & Name & Value & Units & References \\ \hline
    $N$ & Initial tumor size & $10^7-10^9$ & cells  & \citet{del2009does} \\ \hline
    $\lambda_w$ & Wildtype division rate& 0.14 & 1/days  & \citet{bozic2013evolutionary,rew2000cell} \\ \hline
    $\mu_w$ & Wildtype death rate& $0.15-0.21$ & 1/days  & \citet{bozic2013evolutionary} \\ \hline
    $\lambda_a$  & Aneuploid division rate$^\ast$ & 0.14 & 1/days  & - \\ \hline
    $\mu_a$ & Aneuploid death rate$^\ast$ & $0.13-0.21$ & 1/days  & - \\ \hline
    $\lambda_m$ & Mutant division rate& 0.14 & 1/days  & \citet{bozic2013evolutionary,rew2000cell} \\ \hline
    $\mu_m$ & Mutant death rate& 0.13 & 1/days  & \citet{bozic2013evolutionary,carlson2003tumor} \\ \hline
    $u$ & Missegregation rate& $10^{-3}-10^{-2}$ & 1$\slash$cell division  & \citet{nowak2004evolutionary,bakker2023predicting} \\ \hline
    $v$ & Mutation rate& $10^{-5}-10^{-7}$ &  1$\slash$gene$\slash$cell division  & \citet{nowak2004evolutionary,loeb2001mutator} \\ 
    % TODO citing Nowak 2004 for the rates is strange because it is a theory paper - cite the original empirical papers
    \hline
  \end{tabular}
\caption{\textbf{Model parameters.} %NEED DIFFERENT REFS---THESE ARE MOSTLY THEORY PAPERS. CHECK BIONUMBERS? OK
The * symbol in the Name column means that for those parameters the values have not been selected from a paper. Aneuploid birth rate $\lambda_a$ is set to the same value as the wildtype and mutant birth rates, $\lambda_w$ and $\lambda_m$.
Aneuploid death rate $\mu_a$ is set to an intermediate value between the wildtype and mutant death rates, $\mu_w$ and $\mu_m$.}
  \label{table1}
\end{center}
\end{table}

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{Figures/figureAneuploidy.pdf}
\caption{
\textbf{Model illustration.}
\textbf{(A)} A population of cancer cells is composed of wildtype, aneuploid, and mutant cells, which divide with rates $\lambda_w$, $\lambda_a$, and $\lambda_m$ and die at rates $\mu_w$, $\mu_a$, and $\mu_m$, respectively. 
Wildtype cells can become aneuploid at rate $u$. Both aneuploid and wildtype cells can acquire a beneficial mutation with rate $v$. Color denotes the relative growth rates of the three genotypes such that $\lambda_w - \mu_w < \lambda_a - \mu_a < \lambda_m - \mu_m$. \textbf{(B)} The wildtype and the mutant are susceptible and resistant, respectively, to the drug. The aneuploid may be tolerant $(4)$, non-growing $(3)$, partially resistant $(2)$ or fully resistant $(1)$.
}
\label{figureAneuploidy}
\end{figure}

@@

\begin{figure}
\begin{subfigure}{0.5\textwidth}
A\\
\includegraphics[width=1\textwidth]{Figures/ProbvNPlot.pdf}
\end{subfigure}
\\
\begin{subfigure}{0.5\textwidth}
B\\
\includegraphics[width=1\textwidth]{Figures/ThresholdPopulationSizePlot.pdf}
\end{subfigure}
\begin{subfigure}{0.5\textwidth}
C\\
\includegraphics[width=1\textwidth]{Figures/ThresholdPopulationSizeVersusRatioPlot.pdf}
\end{subfigure}
\caption{
\textbf{Aneuploidy facilitates evolutionary rescue of cancer under drug treatment.}
\textbf{(A)} The probability of evolutionary rescue (i.e. the probability that the population does not go to extinction), $\presc$, as a function of the initial tumor size, $N$. Dashed vertical line shows the threshold tumor size, $N_a^*$, above which the probability is very high. Blue dashed line represents the probability of evolutionary rescue as a function of $N$ without aneuploidy ($u=0$). The blue and and red dots represent numerical simulations for the case with aneuploidy and without aneuploidy, respectively. The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the  probability of that a succesful mutant has not ben generated and $n=100$ is the number of simulations.
\textbf{(B)} The threshold tumor size $N_a^*$ as a function of the aneuploid growth rate $\Delta_a$. The dashed horizontal line shows $N^*_m$, the threshold tumor size without aneuploidy ($u=0$). The red dots represent numerical simulations.  The inset highlights the case when aneuploidy cancer cells are non-growing. When aneuploid growth rate is close to or higher than zero, aneuploidy decreases the threshold tumor size, thereby facilitating evolutionary rescue. 
\textbf{(C)} The threshold tumor size $N_a^*$ as a function of the ratio of aneuploidy and mutation rates, $u/v$. The dashed horizontal line shows $N^*_m$, the threshold tumor size without aneuploidy ($u=0$). The blue line represents the exact formula for threshold tumor size $N_a^*$. The red dots represent numerical simulations.  When the aneuploidy rate is much higher than the mutation rate, aneuploidy decreases the threshold tumor size, thereby facilitating evolutionary rescue.
}
\label{rescue_prob}
\end{figure}
%%%%%%%%%%%
\begin{figure}
\vspace*{1\baselineskip}
\includegraphics[width=1\textwidth]{Figures/ReboundProbability.pdf}
\caption{Plot of the probability that a succesful mutant has not appeared by time $t$. The blue line represents the case with aneuploidy ($u>0$), the black line represents the case without aneuploidy ($u=0$).  The green and and red dots represent numerical simulations for the case with aneuploidy and without aneuploidy, respectively.  For the simulations we have chosen the following parameters: $\lambda_w=0.14, \lambda_a=0.14,\lambda_m=0.14,\mu_w=0.17,\mu_a=0.1401,\mu_m=0.13, u=0.14\times10^{-2}, v=0.14\times10^{-7}$. The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the  probability of that a succesful mutant has not ben generated and $n=100$ is the number of simulations. As time increases, aneuploidy plays an important role in helping the cancer cell population escape extinction.}
\label{ReboundProbability}
\end{figure}
%%%%%%%%%%%

@@

%%%%%%%%%%%%
%
%\begin{figure}
%\begin{subfigure}{0.5\textwidth}
%\includegraphics[width=1\textwidth]{Figures/CombinedSubplot.pdf}
%\end{subfigure}
%\begin{subfigure}{0.5\textwidth}
%\includegraphics[width=1\textwidth]{Figures/CombinedSubplotLargePop.pdf}
%\end{subfigure}
%\caption{
%\textbf{Evolutionary rescue probability with partially resistant or tolerant aneuploid cells.}
%Rescue probability is very high when aneuploidy provides partial resistance ($\lambda_a=0.01$), in an initially small tumor (\textbf{A}, $N=10^4$) and even more so in an initially large tumor (\textbf{B}, $N=10^8$). 
%When aneuploidy provides tolerance (\textbf{C}, $N=10^4$; \textbf{D}, $N=10^8$), the rescue probability is much lower. 
%In both scenarios, rescue probability increase with both the wildtype growth rate (x-axis) and the aneuploidy growth rate (colors).
%Markers represent simulation results with 95\% CI; solid and dashed lines for the exact formula (\cref{eq:survival_prob} in \cref{eq:rescue_prob}); dashed lines for the approximate formula (\cref{rescue_prob_approx}), demonstrating that they all agree.
%Parameters: division rate $\lambda_w=\lambda_a=\lambda_m=0.14$ (so that growth rate changes due to variable death rate); mutant death rate $\mu_m=0.13$ (so that mutant growth rate $\Delta_m=0.01$); aneuploidy rate $u=10^{-2}$; mutation rate $v=10^{-7}$.
%}
%\label{rescue_prob_wt_growth}
%\end{figure}
%%%%%%%%%%%%

%\begin{figure}
%% small population size
%% switch from tolerance to non-growing
%% TODO why no switch in panel B?
%% TODO more simulations for 0 > Delta_a > Delta^*_a
%% TODO simulations in A?
%\begin{subfigure}{0.5\textwidth}
%\includegraphics[width=1\textwidth]{Figures/P_est_divergence.pdf}
%\end{subfigure}
%\begin{subfigure}{0.5\textwidth}
%\includegraphics[width=1\textwidth]{Figures/P_est_divergenceLargePopulation.pdf}
%\end{subfigure}
%\caption{\textbf{Evolutionary rescue probability with tolerant or non-growing aneuploid cells.}
%Rescue probability grows with the aneuploid growth rate $\Delta_a$ (x-axis), and is much higher in an initially large tumor than in a small one (\textbf{A}: $N=10^4$; \textbf{B}: $N=10^8$). 
%Markers for simulation results with 95\% CI; 
%solid lines for the exact formula (\cref{eq:survival_prob} in \cref{eq:rescue_prob});
%dashed lines for the approximate formula (\cref{rescue_prob_approx}).
%The approximation agrees with the simulation and exact solution when the initial tumor size is large (panel B).
%When the tumor size is small (panel A), we switch between the approximation for tolerant and for non-growing aneuploid cells; the switch occurs at $1/T^*$. % TODO ref to eq
%Parameters: $\lambda_w=\lambda_a=\lambda_m=0.14$; $\mu_m=0.13$; $u=10^{-2}$; $v=10^{-7}$.
%}
%\label{rescue_prob_an_growth}
%\end{figure}

@@

\section*{Appendix C: Evolutionary rescue time}
% TODO remove the linear approximation, it is not good. OK
